Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-04-07 Manuscript/manuscript.Rmd
manuscript edits

---
title: "Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings"
csl: styles.ref/genomebiology.csl
output:
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
bibliography: References/references.bib
---

@@

**Methods.** We proposed a novel ensemble framework that systematically addresses those data irregularities and characteristics. Multiple random forest classification models, in combination with variations in cell line, data resolution, feature engineering, and re-sampling techniques were performed and compared. A number of performance metrics were assessed, including [??? update when ready].

@@

Multiple random forest classification algorithms were built on domain data extracted from Hi-C data for 2 cell lines $C = \{GM12878, K562\}$, at four different resolutions $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. For each classifier, we compared three predictor types $P = \{\text{OC}, \text{OP}, \text{Distance}\}$ and 4 class balancing techniques $B = \{None, RUS, ROS, SMOTE}$. A system produced from the ensemble framework was defined as: $S = \{c, r, p, b\}$, where $c \in C, \quad R \in R, \quad p \in P,$ and $b \in B$. Letting $|X|$ denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $96 (= 2 \times 4 \times 3 \times 4)$ separate models.

@@

We examined the class imbalance of the different data resolutions for each cell line. As expected, the total number of genomic bins decreased as resolution decreased. That is, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins (Figure 4 (A)). The imbalance ratios, however, remained relatively stable (Figure 4 (B)). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Ideally, a balanced dataset will have an IR close to 0.50. Here, the low imbalance ratios indicated a wide disparity between majority and minority classes (i.e. a genomic bin was much more likely to not overlap with a flanked TAD boundary than it was to overlap with one). Therefore suitable resampling techniques were needed.

@@

By isolating flanked TAD boundaries and, in turn, counting the instances in which annotations overlapped with boundary regions, we were able to assess the correlations of genomic annotations and determine which annotations were enriched together at boundaries.

Clustered correlograms for known colocalized annotations showed that distinct groups of transcription factors were highly positively correlated in both cell lines including CTCF, ZNF143, RAD21, and SMC3ab (Supplementary Figures 2-9). Likewise, combinations of particular histone modifications and chromatin states were also shown to exhibit similar profiles near TAD boundaries including H3K4me1, H3K4me2, and H3K4me3, as well as strong and weak enhancers. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly positively inter-correlated, yet were negatively correlated with many transcription factor binding sites. These results were consistent across resolution and cell line. Notably, correlations amoung annotations were found to increase as resolutions decreased for both cell lines. This indicated that TAD bounardies called, and subsequently flanked, at lower resolutions tended to encapsulate more of these epigenomic marks and contribute to higher levels of enrichment.

The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.

## CLASS IMBALANCE SEVERELY IMPACTS PERFORMANCE OF MODELS WITH OVERLAP COUNTS PREDICTOR TYPES

We first evaluated model performances using only the overlap counts predictor types without re-sampling as these characteristics represented models with the least granularity and could act as a baseline. We measured the performances across resolutions, for each cell line, and found that .

Some notable results became clear. First, we see that the conventional AUC value was unable to accurately portray the poor performance of the models when compared to the composite metrics F1-score, MCC, and, to a lesser extent, the balanced accuracy. Inspecting the F1-score and MCC more closely, we see that some models at lower resolutions (for either cell line) were unable to correctly predict any bins that overlap with flanked TAD boundaries in the test data, as indicated by the missing bars.

## DATA RE-SAMPLING IMPROVES MODEL PERFORMANCE

Comparisons of the effects of the various resampling techniques were then made across resolutions, for each cell line. Results showed that performances drastically improved among the composite metrics (Figure 8 (B-D)). Models implemented on randomly under-sampled and SMOTE-sampled data performed similarly, while performances on randomly over-sampled data lagged behind, yet still outperformed models with no data re-sampling. Overall, there appeared the be a downward trend in model performance as resolutions decreased, across both cell lines.

@@

After illustrating the improvement in model performance using re-sampled data for only the overlap counts predictor types, we then directly compared the performances of each re-sampling techniques on the other two predictor types (overlap percents and distances). Immediately, results showed that models with distance-type predictors outperformed models with other predictor types (Figure 8 (A-D)). This results was consistent across resolution and cell line. As before, we are able to gain much more insight into the performances of the models by using composite metrics as opposed to simply the AUC, in cases of class imbalance. Likewise, performances were shown to continue to degrade as resolutions decreased.

## MODELS IMPLEMENTED ON SMOTE RE-SAMPLED DATA WITH DISTANCE-TYPE PREDICTORS PERFORMED BEST AMOUNG ALL ENSEMBLE SYSTEMS

Of the re-sampling techniques considered, it was found that SMOTE re-sampled datasets, implemented on distance-type predictors, yielded the highest model performances across each of the 3 composite metrics (Figure 9). These results were consistent across resolution and cell line. Although RUS performed only marginally worse than SMOTE, there was a steap drop off in performances for ROS and no re-sampling, across both resolution and cell line. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do.

Furthermore, we found that model performances tended to decline as data resolution decreased. Random forest models implemented on 10 kb, SMOTE resampled, data resolutions, with distance-type predictors were found to perform best according to MCC and balanced accuracy. There was a slight improvement in F1-Score for the 25 kb resolution on the GM12878 cell line. Performances from the K562 cell line degraded much faster as resolution decreased, while the worst performing models for the GM12878 cell line were at the 50 kb resolution, followed by a slight improvement in performance at the 100 kb resolution.

>>>>

---
title: Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance
  Settings
csl: styles.ref/genomebiology.csl
output:
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
  pdf_document: default
bibliography: References/references.bib
---

@@

**Methods.** We proposed a novel ensemble framework that systematically addresses those data irregularities and characteristics. Multiple random forest classification models, in combination with variations in cell line, data resolution, feature engineering, and re-sampling techniques were performed and compared. A number of performance metrics were assessed, including [??? will update when ready].

@@

Multiple random forest classification algorithms were built on domain data extracted from Hi-C data for 2 cell lines $C = \{GM12878, K562\}$, at four different resolutions $R = \{\text{10 kb}, \text{25 kb}, \text{50 kb}, \text{100 kb\}}$. For each classifier, we compared three predictor types $P = \{\text{OC}, \text{OP}, \text{Distance}\}$ and 4 class balancing techniques $B = \{None, RUS, ROS, SMOTE\}$. A system produced from the ensemble framework was defined as: $S = \{c, r, p, b\}$, where $c \in C$, $r \in R$, $p \in P$, and $b \in B$. Letting $|X|$ denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $96 (= 2 \times 4 \times 3 \times 4)$ separate models.

@@

We examined the class imbalance of the different data resolutions for each cell line. As expected, the total number of genomic bins decreased as resolution decreased (Figure 4A). That is, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins. The imbalance ratios, however, remained relatively stable across resolutions (Figure 4B). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Ideally, a balanced dataset will have an IR close to 0.50. Here, the low imbalance ratios indicated a wide disparity between majority and

@@

By isolating flanked TAD boundaries and, in turn, counting the instances in which annotations overlapped with boundary regions, we were able to assess the correlations of genomic annotations and determine which annotations were enriched together at boundaries. Clustered correlograms for known colocalized annotations showed that distinct groups of transcription factors were highly positively correlated in both cell lines, and across resolutions, including CTCF, ZNF143, RAD21, and SMC3ab (Supplementary Figures 2-9). Likewise, combinations of particular histone modifications and chromatin states were also shown to exhibit similar profiles near TAD boundaries including H3K4me1, H3K4me2, and H3K4me3, as well as strong and weak enhancers. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly positively inter-correlated, yet were negatively correlated with many transcription factor binding sites. Notably, correlations among annotations were found to increase as resolutions decreased for both cell lines. This indicated that TAD bounardies called, and subsequently flanked, at lower resolutions tended to encapsulate more of these epigenomic marks and contribute to higher levels of enrichment.

<!--
The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation. 
-->

## CLASS IMBALANCE SEVERELY IMPACTS PERFORMANCE AMONG ALL CLASSES OF MODELS

We first evaluated performances for models implemented using overlap counts predictor types, without re-sampling (on the GM12878 cell line), as these characteristics represented models with the least granularity and could act as a baseline. In evaluating such composite metrics like F1-Score and MCC, we found that these models exhibited poor performances across resolutions (Figure 5A). Furthermore, at lower resolutions (50 kb and 100 kb specifically), the F1-Score and MCC could not be determined given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives, as can be seen by the missing red points. When including overlap percents perdictor types in the models, coupled with no resampling, we found that performances were worse than when using OC predictor types (Figure 5B). As with OC predictor types, these models were also not able to yield F1-Scores and MCCs at the 50 kb and 100 kb resolution. Models built on distance-type predictors, however, exhibited slightly better performances than the other two types, yet still yielded missing or low F1-Scores and MCC at lower resolutions (Figure 5C). This indicated that the key driver leading to poor performances was not feature engineering, but instead the severity in the class imbalance.

## DATA RE-SAMPLING IMPROVED PERFORMANCE AMONG ALL CLASSES OF MODELS

Comparisons of the effects of the various resampling techniques were then made. We found that there was a substantial increase in F1-Score and MCC for models built on resampled data using any of the 3 techniques (Figure 5). After resampling, we were able to report performances at the lower resolutions which was not feasible for models built on fully imbalanced data. For all predictor types, implementation of either RUS or SMOTE resampling was found to yield consistently better results than no resampling and ROS.

@@

We then assessed the performances of models implemented on distance type predictors alone. Results showed that models with distance-type predictors outperformed models with other predictor types (Figure 5C). This result was consistent across both resampling technique and resolution. Directly comparing model performances using distance type predictors, we found that SMOTE resampled datasets yielded the highest model performances across each of the composite metrics. These results were consistent across resolution and cell line. Although RUS performed only marginally worse than SMOTE, there was a steap drop off in performances for ROS and no re-sampling, across both resolution and cell line. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do.

Furthermore, we found that when using distance-type predictors, model performances tended to decline as data resolution decreased. There was a slight improvement in F1-Score for the 25 kb resolution on the GM12878 cell line, followed by a steep decline at 50 kb. Both F1-Scores and MCCs showed a momentary increase in performance at the lowest resolution considered in the study. Performances from the K562 cell line degraded much faster as resolution decreased, while the worst performing models for the GM12878 cell line were at the 50 kb resolution, followed by a slight improvement in performance at the 100 kb resolution.
